Numerical Investigation of Multi-Fluid Flows through Dual Saturated-Unsaturated Porous Media

### 2015-11-06 Papers/ViscousInstabilities/Manuscript_ViscousInstabilities_KC.tex
Update Manuscript_ViscousInstabilities_KC.tex

For reservoir engineers and hydrogeologists, mapping \red{(you don't map a behaviour)} the behaviour of multiphase flows and predict the response of the system are major challenges followed by complex geometries with irregular boundaries and heterogeneous geological formations \red{(sentence needs to be improved)}. These are some of the main challenges \red{(too many challenges in the same paragraph ...)} for accurately forecasting the behaviour of flows into porous media. Both permeability and porosity are important parameters in the equations used for numerical simulation of multiphase flow in porous media.\red{(what was the objective of this sentence?)}  %Heterogeniety of permeability and porosity is a product of the medium's formation process %-- deposition and diagenesis -- and exists at different scales. 
At the microscopic level, the continuum approach can be applied to a representative elementary Volume (REV) to define these parameters~\citep{bastian_phd1999}. This is well suited for analysis of small scales such as core samples \red{(These sentences are not linked with the previous sentences, also what is the objective of them? )}.

Although parallel processing is available nowadays, a fine-scale geological model is far too detailed to be accumulated by common reservoir simulators, mainly because of the processing time. \red{(Same comment as above!)}

Therefore larger scales can be approached by upscaling techniques. This technique which is usually employed to solve the problem fails for certain types of heterogeneity. These cases are normally resolved by stochastic representation of permeability and porosity fields. 
%Various upscaling techniques, including analytical and numerical, have been introduced in the literature to calculate upscaled flow properties.
There are several approaches to upscale the solutions, while three common types are herein discussed. The $1^{st}$ assumes capillary equilibrium and calculates effective relative permeability using a suitable technique, such as numerical methods, percolation methods and space renormalization methods. The $2^{nd}$ approach is called pseudoization. It generates a psuedo relative permeability by matching immiscible displacement simulation of a coarse-grid to the output of fine-grid simulations. The $3^{rd}$ approach involves direct calculation of relative permeability from displacement experiments.~\citep{johnson_1959, jones_1978}.

The first approach is limited to low flow rates and very small heterogeneity. The second approach overcomes these limitation but also posses drawbacks which include; availability of fine-grid simulation with suitable level of accuracy, dependence of the psuedo functions on rate, direction and grid specification. The experimental data from the third approach is only suitable for a small class of scenarios.~\citep{Chang_1994}. While  non-uniform permeability and porosity may be easy to deal with at a microscopic scale for porous media flow equations, it presents a lot of challenges at a meso/field scale, which is required for reservoir simulations. \red{From `At the microscopic level ...' to `...reservoir simulations' --try to rewrite these sentences/paragraphs ... they are very confusing and not really connected'}

@@

%\medskip
Under certain conditions the displacement of fluids in porous media is unstable \red{(what sort of conditions?)} and results in the formation of strange patterns \red{(sentence is not very technical} in the form of \textit{fingers}. This phenomenon is referred to as \textit{viscous fingering}. The flow of an injected fluid into a geological formation (porous medium) generates instabilities due to the heterogeneities of the porous medium \red{(Is this correct? Is it your opnion? If so built up the facts to substantiate your conclusions. If it is not your opinion then give reference(s))}. Even though the mechanism of the instabilities may vary (in immiscible fluids the most important factor is the surface tension while in miscible fluids, it is the diffusion \red{(thermal or mass diffusion ...be specific)}), however it is still a function of several variables such as, relative mobility, gravity, dispersion, permeability differences. Moreover, in most cases interfacial tension is directly linked with the mobility ratio (or viscosity ratio) between the displaced and the displacing fluid.

Both the evolution of the computers and the advances in modelling and simulation of flows in porous media may help to visualise and address the importance of this phenomenon which can be found in many petroleum, chemical and environmental related applications. \red{(sentence is not written in a technical format, also what is its objective?)}

%\medskip
\begin{comment}
Finite difference methods (FDM) have been extensively used in most industry-standard flow simulators~\citep{aziz_1986, chen_2005, chang_1990} with relative success in multi-physics single phase applications (e.g., black oil model, passive tracer advection etc). However, FDM are often limited to relatively simple geometries (with low grid quality and orientation) representing idealised geological formations~\citep{mlacnik_2004, king_1999}. Additionally, FDM produce excessive numerical dispersion when heterogeneity (represented by permeability and porosity fields) is present~\citep{chavent_1986}. The geometrical flexibility associated with high-order numerical accuracy of finite element methods (FEM) has proven to overcome these deficiencies.

@@

\citet{cumming_2011} demonstrated that CVFEM discretisation could also be used to solve Richards' equations (coupled mass conservation and Darcy equation) in heterogeneous media with relatively small computational overhead~\citep[compared with traditional coupled velocity-pressure based formulations, see also][]{cumming_phd2012}. Fluxes over CVs were calculated based upon material properties, whereas the saturation fields were volume-averaged at the interface of the materials, enforcing mass balance as described by \citet{kirkland_1992}.
\end{comment}

@@

In this paper, the overlapping control volume finite element method (OCVFEM), formulation recently introduced by~\citet{jackson_2013}~\citep[see also][]{gomes_2013,pavlidis_2013} is used to investigate the development of multiphase flows through heterogeneous porous media \red{(what is the objective of this paper? To study multiphase porous media flows? Or to study a particular physical aspect of multiphase flows in porous media? You need to be precise here! )}. In the OCVFEM, a dual pressure-velocity representation in CV and FEM spaces  are embedded into a family of triangular/tetrahedral element pairs, P$_{m}$DG-P$_{n}$~\citep{cotter_2009b} and P$_{m}$DG-P$_{n}DG$~\citep{xie_2014}. In these element types, the dual velocity and pressure fields are represented by $m^{\rm th}$- (discontinuous) and $({n})^{\rm th}$-order (continuous/discontinuous) polynomials, respectively.
Moreover, velocity field has dual mapping in FEM and CV spaces. This approach can achieve an exact representation of the Darcy force balance equations.\textcolor{red}{The main objective of this paper is to investigate the impact of heterogeneous porous media in fingers formation using a FEM-based unstructured fixed mesh multi-fluid model with uniform permeability distribution under both saturated and unsaturated porous media}. 
In addition, the study focused in flow instabilities determined by the viscosity ratio. In the simulations presented in this paper the element pairs as described above and the overlapping control volume finite element method (CVFEM) model were used, in which the dual velocity-pressure fields are embedded in FEM space with simultaneous projection into CV space. Scalar fields (e.g., saturation, concentration, density, etc) are represented in the control volume space. High-order accurate downwind schemes on element boundaries on discontinuous scalar fields are flux-limited (based on NVD approach) to obtain bounded and compressive (capturing the interfaces) solutions. Section $2$ summarises the numerical formulation used in this work, followed by description of the flow instabilities  and a description of model validation. Finally, conclusions are drawn in the final section.

@@

During the simulations performed in this paper the Viscous fingering phenomenon has been observed. Viscous fingering  is highly related with the mobility, which describes how easily a fluid moves into a porous media and can be expressed through the mobility ratio (\textit{MR}). The ratio between the mobility of the displacing phase to the mobility of the displaced.

>>>>

Reservoir engineers and hydrogeologists are dealing with complex geometries, irregular boundaries and heterogeneous geological formations. Forecasting the behaviour of multiphase flows or the future response of the system are major challenges that need to be taken into account. 
%Both permeability and porosity are important parameters in the equations used for numerical simulation of multiphase flow in porous media.\red{(what was the objective of this sentence?)}  %Heterogeniety of permeability and porosity is a product of the medium's formation process %-- deposition and diagenesis -- and exists at different scales. 
To tackle these problems the continuum approach, in a microscopic level, can be applied to a representative elementary Volume (REV) to describe the spatial variation in parameters such as the porosity and permeability~\citep{bastian_phd1999}. %This is well suited for analysis of small scales such as core samples. \red{(These sentences are not linked with the previous sentences, also what is the objective of them? )}.

Larger scale problems can be approached by upscaling techniques. This technique which is usually employed to solve the problem fails for certain types of heterogeneity. These cases are normally resolved by stochastic representation of permeability and porosity fields. 
%Various upscaling techniques, including analytical and numerical, have been introduced in the literature to calculate upscaled flow properties.
There are several approaches to upscale the solutions, while three common types are herein discussed. The $1^{st}$ assumes capillary equilibrium and calculates effective relative permeability using a suitable technique, such as numerical methods, percolation methods and space renormalization methods. The $2^{nd}$ approach is called pseudoization. It generates a psuedo relative permeability by matching immiscible displacement simulation of a coarse-grid to the output of fine-grid simulations. The $3^{rd}$ approach involves direct calculation of relative permeability from displacement experiments.~\citep{johnson_1959, jones_1978}.

The first approach is limited to low flow rates and very small heterogeneity. The second approach overcomes these limitation but also posses drawbacks which include; availability of fine-grid simulation with suitable level of accuracy, dependence of the psuedo functions on rate, direction and grid specification. The experimental data from the third approach is only suitable for a small class of scenarios.~\citep{Chang_1994}. While  non-uniform permeability and porosity may be easy to deal with at a microscopic scale for porous media flow equations, it presents a lot of challenges at a meso/field scale, which is required for reservoir simulations.

@@

%\medskip
Under certain \red{(spatial or geological, which is the best word?)} conditions the displacement of fluids in porous media is unstable and results in the formation of irregular flow patterns in the form of \textit{fingers}. This phenomenon, that describes these flow instabilities mentioned above is referred to as \textit{viscous fingering}. Even though the mechanism of the instabilities may vary (in immiscible fluids the most important factor is the surface tension while in miscible fluids, it is the diffusion \red{(thermal or mass diffusion ...be specific)}), however it is still a function of several variables such as, relative mobility, gravity, dispersion, permeability differences. Moreover, in most cases interfacial tension is directly linked with the mobility ratio (or viscosity ratio) between the displaced and the displacing fluid.

Both the evolution of visualisation techniques through the extensive use of parallel computing as well as the advances in modelling and simulation of flows in porous media may help to represent and address the importance of this phenomenon which can be found in petroleum, chemical and environmental related applications.

%\medskip
%\begin{comment}
Finite difference methods (FDM) have been extensively used in most industry-standard flow simulators~\citep{aziz_1986, chen_2005, chang_1990} with relative success in multi-physics single phase applications (e.g., black oil model, passive tracer advection etc). However, FDM are often limited to relatively simple geometries (with low grid quality and orientation) representing idealised geological formations~\citep{mlacnik_2004, king_1999}. Additionally, FDM produce excessive numerical dispersion when heterogeneity (represented by permeability and porosity fields) is present~\citep{chavent_1986}. The geometrical flexibility associated with high-order numerical accuracy of finite element methods (FEM) has proven to overcome these deficiencies.

@@

\citet{cumming_2011} demonstrated that CVFEM discretisation could also be used to solve Richards' equations (coupled mass conservation and Darcy equation) in heterogeneous media with relatively small computational overhead~\citep[compared with traditional coupled velocity-pressure based formulations, see also][]{cumming_phd2012}. Fluxes over CVs were calculated based upon material properties, whereas the saturation fields were volume-averaged at the interface of the materials, enforcing mass balance as described by \citet{kirkland_1992}.
%\end{comment}

@@

In this paper, \textcolor{red}{the main objective is to investigate the impact of heterogeneous porous media in fingers formation using a FEM-based unstructured fixed mesh multi-fluid model with uniform permeability distribution under both saturated and unsaturated porous media}. The overlapping control volume finite element method (OCVFEM), formulation recently introduced by~\citet{jackson_2013}~\citep[see also][]{gomes_2013,pavlidis_2013} is used to investigate the development of multiphase flows through heterogeneous porous media.This finite element based method has a dual pressure-velocity representation in CV and FEM spaces  are embedded into a family of triangular/tetrahedral element pairs, P$_{m}$DG-P$_{n}$~\citep{cotter_2009b} and P$_{m}$DG-P$_{n}DG$~\citep{xie_2014}. The dual velocity and pressure fields are represented by $m^{\rm th}$- (discontinuous) and $({n})^{\rm th}$-order (continuous/discontinuous) polynomials, respectively.
Moreover, velocity field has dual mapping in FEM and CV spaces. This approach can achieve an exact representation of the Darcy force balance equations. 
In addition, the study focused in flow instabilities determined by the viscosity ratio. In the simulations presented in this paper the element pairs as described above and the overlapping control volume finite element method (CVFEM) model were used, in which the dual velocity-pressure fields are embedded in FEM space with simultaneous projection into CV space. Scalar fields (e.g., saturation, concentration, density, etc) are represented in the control volume space. High-order accurate downwind schemes on element boundaries on discontinuous scalar fields are flux-limited (based on NVD approach) to obtain bounded and compressive (capturing the interfaces) solutions. Section $2$ summarises the numerical formulation used in this work, followed by description of the flow instabilities  and a description of model validation. Finally, conclusions are drawn in the final section.

@@

During the simulations performed the Viscous fingering phenomenon has been observed. Viscous fingering  is highly related with the mobility, which describes how easily a fluid moves into a porous media and can be expressed through the mobility ratio (\textit{MR}). The ratio between the mobility of the displacing phase to the mobility of the displaced.
